---
title: "Activity Monitoring Data Analysis"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and Preprocessing the Data

```{r load-data}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Load the dataset
activity_data <- read.csv("activity.csv")

# Convert date column to Date format
activity_data$date <- as.Date(activity_data$date)

# View first few rows
head(activity_data)
```

## What is the mean total number of steps taken per day?

```{r total-steps-per-day}
# Compute total steps per day, ignoring NA values
total_steps_per_day <- activity_data %>% group_by(date) %>% summarize(total_steps = sum(steps, na.rm = TRUE))

# Plot histogram
ggplot(total_steps_per_day, aes(x = total_steps)) +
  geom_histogram(binwidth = 1000, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Total Steps per Day", x = "Total Steps", y = "Frequency")

# Calculate mean and median
mean_steps <- mean(total_steps_per_day$total_steps, na.rm = TRUE)
median_steps <- median(total_steps_per_day$total_steps, na.rm = TRUE)

# Print values
mean_steps
median_steps
```

## What is the average daily activity pattern?

```{r activity-pattern}
# Compute mean steps per 5-minute interval
average_steps_per_interval <- activity_data %>% group_by(interval) %>% summarize(mean_steps = mean(steps, na.rm = TRUE))

# Time series plot
ggplot(average_steps_per_interval, aes(x = interval, y = mean_steps)) +
  geom_line(color = "red") +
  labs(title = "Average Daily Activity Pattern", x = "5-minute Interval", y = "Average Steps")

# Find the interval with the max number of steps
max_interval <- average_steps_per_interval$interval[which.max(average_steps_per_interval$mean_steps)]
max_interval
```

## Imputing Missing Values

```{r impute-missing-values}
# Count total NA values
missing_values <- sum(is.na(activity_data$steps))

# Replace NA values with the mean of the respective 5-minute interval
activity_data_filled <- activity_data
for (i in 1:nrow(activity_data_filled)) {
  if (is.na(activity_data_filled$steps[i])) {
    activity_data_filled$steps[i] <- average_steps_per_interval$mean_steps[average_steps_per_interval$interval == activity_data_filled$interval[i]]
  }
}

# Compute total steps per day with imputed values
total_steps_per_day_filled <- activity_data_filled %>% group_by(date) %>% summarize(total_steps = sum(steps))

# Plot histogram after imputing missing values
ggplot(total_steps_per_day_filled, aes(x = total_steps)) +
  geom_histogram(binwidth = 1000, fill = "green", color = "black", alpha = 0.7) +
  labs(title = "Total Steps per Day (After Imputation)", x = "Total Steps", y = "Frequency")

# Compute new mean and median
mean_steps_filled <- mean(total_steps_per_day_filled$total_steps)
median_steps_filled <- median(total_steps_per_day_filled$total_steps)

# Print values
mean_steps_filled
median_steps_filled
```

## Are there differences in activity patterns between weekdays and weekends?

```{r weekday-weekend-patterns}
# Create a new variable for weekday or weekend
activity_data_filled$day_type <- ifelse(weekdays(activity_data_filled$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")

# Compute mean steps per interval for weekdays and weekends
average_steps_by_day_type <- activity_data_filled %>% group_by(interval, day_type) %>% summarize(mean_steps = mean(steps))

# Plot time series by weekday and weekend
ggplot(average_steps_by_day_type, aes(x = interval, y = mean_steps, color = day_type)) +
  geom_line() +
  facet_wrap(~day_type, ncol = 1) +
  labs(title = "Activity Patterns: Weekdays vs. Weekends", x = "5-minute Interval", y = "Average Steps", color = "Day Type")
```

## Conclusion

- The dataset contains missing values which were imputed using the mean of each 5-minute interval.
- The activity patterns vary between weekdays and weekends, with different peak times.
- The new mean and median after imputing missing values are slightly different from the original values.
